---
title : "<center>__AsaLab__</center><center>A package for filtering and analysis CT values </center>"
author: "<center>*Åsa Torinsson Naluai, Mohammad Tanvir Ahamed*</center><center>University of Gothenburg, Sweden</center>"
date: "<center> Date: *`r Sys.Date()`*</center><center>__*Version: 1.0.0*__</center>"

output:
  BiocStyle::html_document: 
    toc: yes
    toc_depth: 4
    number_sections: TRUE
    
  BiocStyle::pdf_document: 
    toc: TRUE
    toc_depth: 4
    number_sections: TRUE
  
vignette: >
  %\VignetteIndexEntry{AsaLab User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8x]{inputenc}

---
AsaLab is a package for filtering, analysing and visualization of CT valuse. 

# Data formate for input file 
## Data formate for CT values
Input data is the output of "ExpressionSuite software" in .txt formated file. The data structure is as follows

```{r, echo=FALSE}
library(AsaLab)
data(ex_data_1)
head(ex_data_1)
```

## Data formate for phenotype values
Input data formate for phenotype data is as follows. The first column is the sample name and the second column is the phenotype / status for the corresponding sample. 
```{r, echo=FALSE}
library(AsaLab)
data(ex_phenotype_1)
head(ex_phenotype_1)
```
# Function to read files (CT values and Phenotype)
## Read CT values form ExpressionSuite output
The function named `readCT` can read the file for CT values for an external file in csv, xls or txt formate.
```r
mdata <- readCT(file, skip = 14,...)

# Explanation of the function parameters
# -- file     =  Name of the file to read including location
# -- skip     =  Number of line to skip form reading. Default is 14.
# -- ...      =  additional parameter to input.
```
## Read phenotype file
The phenotype file can be read by the following code. 
```r
phen   <- read.csv (file , sep="\t")

# Explanation of the function parameters
# -- file     =  Name of the file to read including location
```

# Filtering CT values
## CT value filtering criteria
Based on the data formate in the previous section, there are some criteria to filter the CT values. The criteria are as follows

- In 'Omit' column if values are TRUE, delete those row/sample.
- In 'Ct' column replace all values labeled "Undetermined" by NA or 40 (User specified).
- In 'Ct' column any value beyond 15 and 40 , replace by NA
- __Sample with missing values:__
-  1. *Sample has 1 non-missing value*: Replace all missing with that one non-missing value.
-  2. *Sample has 2 non-missing value*: d is Absolute difference between two CT values. If d > 2 (User specified), delete all those sample/row and if d <= 2 (User specified), replace all missing values by the mean of 2 non-missing CT values. 
-  3. *Sample has 3 non-missing value*: Sort the 3 non-missing CT values by increasing order. d1 ( difference between 1st and 2nd) and d2 (difference between 2nd and 3rd) is the consicutive absolute difference between them. If any d (d1 or d2) is greater than 2 (User specified), delete all sample/ row. If d2 > 2d1, delete the 3rd CT value and replace all the non-missing ct values with the mean of 2 non-missing values. If d1 > 2d2, delete the 1st CT value and replace all the non-missing ct values with the mean of 2 non-missing values
-  4. *Sample has 4 or more  non-missing value*: Count the median of all non missing values and delete those, whose absolute difference than median is greater than 2 (User specified). Then count the mean of the rest and  replace the missing values with mean.
- __Sample without missing values:__
-  1. *Sample with 2 values* : If d > 2 (User specified), delete both of samples otherwise keep both.
-  2. *Sample with 3 values* : Sort the 3 CT values by increasing order. d1 ( difference between 1st and 2nd) and d2 (difference between 2nd and 3rd) is the consicutive absolute difference between them. If any d (d1 or d2) is greater than 2 (User specified), delete all sample/ row. If d2 > 2d1, delete the 3rd CT value and if d1 > 2d2, delete the 1st CT value.
-  3. *Sample with 4 or more values* : Count the median of ct values and delete those, whose absolute difference than median is greater than 2 (User specified) and keep the rest.

## Applying CT value filtering criteria
The above criteries has applied with the function `filterCT` and  mean CT value has computed for each sample for specific genes and save them in a file.
```r
res <- filterCT(file, sample = "Sample.Name",
                     target = "Target.Name", ct = "Ct",
                     ctlimit = c(15,40), ampscore = 0.6,
                     undet = NULL, phenotype = NULL,ctd = 2,
                     omit = NULL, out = "filterCT")
                     
# Explanation of the function parameters
#  -- file      = Object output of \link[AsaLab]{readCT}.
#  -- sample    = Sample column name. Default "Sample.Name"
#  -- target    = Target column name. Default "Target.Name"
#  -- ct        = CT column name. Default "Ct"
#  -- ctlimit   = Limit to keep CT value. Default min= 15 and max = 40.
#  -- undet     = Value to replace undetermine. Default is NULL. Input value should be numeric.
#  -- ampscore  = Amplification score to reject the sample. Default 0.6 [Not active in the current version]
#  -- phenotype = Group classification for sample name.
#  -- ctd       = Absulute difference between ct values to accept. Default is 2. See user documentation for details.
#  -- omit      = If "TRUE" , sample with TRUE value in omit column will delete. Default is NULL.
#  -- out       = Name of output file to save the result in CSV formate. Default name "filterCT".
```
### Examples
```{r}
## CT values form ExpressionSuite output
data(ex_data_1)
mdata <- ex_data_1
## Phenotype file
data(ex_phenotype_1)
phen <- ex_phenotype_1

## Apply CT value filtering criteria
# Without phenotype file input
res  <- filterCT(file = mdata, phenotype = NULL, out = "filterCT")
head(res)
# With phenotype file input
res  <- filterCT(file = mdata, phenotype = phen, out = "filterCT")
head(res)
```
The above output will be saved in the working directoty in a file (Default file name is "filterCT.csv". File name can be change in "out" parameter of `filterCT` function)

# Analyze CT values
## nalysis to apply
The following steps has applied to analyze filtered CT values (With `filterCT` function)

- Delete both gene (column) and sample (row) with missing values. Thrashold is defined by user. Default is 70%. That means, column (Gene) with equel or more than 70% missing value will be exculded form the analysis. Then the same approach will apply over row (Sample).
- The rest of the missinf values will be replace by user define valus or not. Default is NULL.
- Save this result in the working directory in a file named "filterCT_1.csv".
- Based on the list of housekeeping gene, deltaCT is computed and saved in the working directory in a file named as corresponding housekeeping gene in csv formate.

---- Based on a single housekeeping gene, boxplot and p-values (Based on 2 sample t-test) of deltaCT values for different phenotype group has computed and saved in the working directory in a file named as corresponding housekeeping gene in pdf formate.

- Based on the list of housekeeping gene, the average CT values for those housekeeping gene is computed. Then deltaCT is computed and saved in the working directory in a file named as "avg_deltaCT.csv".

---- Based on a single housekeeping gene, boxplot and p-values (Based on 2 sample t-test) of deltaCT values for different phenotype group has computed and saved in the working directory in a file named as corresponding housekeeping gene in pdf formate. 

The above steps has applied with `analyzeCT` function
```r
CTres <- analyzeCT(file, skip = 2, hkgene ,control="CONTROL", del = 0.7, missCT = NULL)

# Explanation of the function parameters
- file      = Data in Matrix. Output of `filterCT` function.
- skip      = Number of column to skip form 1st column in input file. Default is 2.
- hkgene    = House keeping gene.
- control   = Name of phynotype that will assign as control. Default is "CONTROL".
- del       = Percentage of missing value for which any gene or sample will be excuded form analysis. Default is 0.7 (70 percentage).
- missCT    = If missing value will replace with 40 or otheres. Default is NULL.
```
## Examples
```{r, message=FALSE}
## CT values form ExpressionSuite output
data(ex_data_1)
mdata <- ex_data_1
## Phenotype file
data(ex_phenotype_1)
phen <- ex_phenotype_1
## Filter CT values 
res  <- filterCT(file = mdata, phenotype = phen, out = "filterCT")
## House keeping gene
hkg    <- c("YWHAZ_047","GUSB_627","HPRT1_909")
CTres  <- analyzeCT (file = res, skip=2, hkgene = hkg,control="control",del = 0.7, missCT = NULL)
```

The output of `analyzeCT` has 3 (Three) section/ part

```r
# Part 1 : Deleted gene list
p1 <- CTres$del_gene
p1
```
```{r, echo=FALSE}
p1 <- CTres$del_gene
p1
```

```r
# Part 2 : Deleted sample list
p2 <- CTres$del_sample
p2
```
```{r, echo=FALSE}
p2 <- CTres$del_sample
p2
```

```r
# Part 3 : Delta CT values based on corresponding housekeeping genes
p3<- lapply(CTres$delta_CT,head)
p3
```
```{r, echo=FALSE}
p3<- lapply(CTres$delta_CT,head)
p3
```
